---
title: "exercice centrage reduction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
experts <- read.table(file="data/perfumes_qda_experts.csv",header=TRUE, sep=",",quote="\"")
for (col in colnames(experts[,1:4])){
  experts[, col] <- as.factor(experts[, col])
}
```

# Center and reduce 

```{r}
library(SensoMineR)
library(dplyr)
```

We want to center and reduce the matrix of products. Let's choose 7 products and keep 8 sensory attributes:
```{r eval=FALSE}
df <- ____
```

```{r}
#Soluce
df <- experts %>%
  select(c(Panelist,Product, Floral, Marine, Fruity, Heady, Wrapping, Oriental, Greedy, Vanilla)) %>%
  filter(Product == "Angel" | Product == "Pleasures" | Product == "J'adore EP" | Product == "Aromatics Elixir" | Product == "Shalimar" | Product == "Chanel N5" | Product == "Lolita Lempicka")
```

Next, we drop the levels of factors with the `droplevels` function:
```{r}
df <- droplevels(df)
```

#First step : center

Centering a matrix is do the difference with the means of each columns. 

Using the `averagetable` function from `SensoMineR`, we can get the matrix of means for adjusted means. We're gonna center this matrix.
```{r}
resaverage <- averagetable(df, formul = "~Product+Panelist", firstvar = 3)
res <- as.matrix(resaverage)
res
```
To center it, we want to do the difference between each adjusted means with the mean of the sensory attribute corresponding to the column. 

To do it, we must to obtain the means of each sensory attributes. The first point is to sum values for each columns:

```{r}
I <- nrow(res)
J <- ncol(res)
g <- rep(1,I)%*%res
g
```
and next we weight them by $n$ (here $n=7$); according the definition of mean: 

```{r}
D <- diag(1/I, J)
g <- g%*%D
g
```
This vector is the center of gravity of the attribute's space. Next, we must to get a matrix of dimensions $(p,q)$ (where $p$ is the number of products and $q$ is the number of sensory descriptors) of means for each column.

```{r}
col_1 <- matrix(1,I,1)
col_1%*%g
```

The last step to centering is to do the difference:
```{r}
res_c <- res-col_1%*%g
```

# Second step : reduce

To reduce the matrix of adjusted and centered means, we go to do the matrix product with $\frac{1}{\sigma}$. Let's get the vector of variances of each sensory attributes to begin:
```{r}
var <- diag(t(res_c)%*%res_c)/I
```

Next we calculate $\sigma$ and we get the matrix of $\frac{1}{\sigma}$ of dimensions $(p,q)$:
```{r}
inv_sigma <- 1/sqrt(var)
inv_sigma

m_sigma <- matrix(diag(1/sqrt(diag(t(res_c)%*%res_c)/I)),ncol=J)
m_sigma
```

to finish, we do the product:
```{r}
res_cr <- res_c%*%m_sigma
```

A good way to check if a matrix is well centered reduced is to calculate if to verify if standards deviations are equal to 1:
```{r}
sqrt(diag(t(res_cr)%*%res_cr)/I)
```

