---
title: "Quanti_01"
author: "Guillaume"
date: "01/06/2022"
output: html_document
---

```{r}
perfumes <- read.csv("data/perfumes_qda_experts.csv")
```

## PART 1A
The empirical distribution of a variable is the set of values (or modalities) taken by this variable, and their associated numbers. Here we want to have an idea of the distribution of the different values of the series associated with the variables `vanilla` and `Floral`.

however, the variables being continuous, it is very rare to have the same value twice for two different individuals.

We therefore choose to make a grouping in classes.

Let's illustrate these notions with the `his` function:
```{r}
hist(perfumes$Vanilla)
```
Here, a lot of information is available to us. The first one is an idea of the representation of the scores given by the judges to the attribute `vanilla`. We that a lot of notes are between 0 and 1. The rest is more homogeneous. As explain previously, this distribution is cut in classes. The cut is set by the argument `breaks`. By default, it is set with the Sturges method, which is the most recommended in the most of the cases. If you specify manually this number,  make sure the number is not too high or not too low.

```{r}
hist(perfumes$Vanilla, breaks = 50)
```
We notice however that this histogram represents the frequencies of observations on the ordinate, or more precisely the number of observations of each class. It is the argument `freq = True` that gives us this representation. But you sometimes histograms are plot using Frequency density, the frequency per unit for the data in each class. It is calculated by dividing the frequency by the class width. We change the argument `probability` to `probability = TRUE` to have this representation.

```{r}
hist(perfumes$Vanilla, breaks = 50, probability = TRUE)
```

## PART 1B

This notion of density is introduce here, it allows to restrict the frequencies of the classes between 0 and 1.
We can use the `density` function to estimate the density of a quantitative variable. 
```{r}
plot(density(perfumes$Vanilla))
```
The histogram of frenquency density can be superposed to the curve :
```{r}
hist(perfumes$Vanilla, breaks = 50, probability = TRUE)
lines(density(perfumes$Vanilla), col="red")
```
we can also compare two estimate density of two different variables:
```{r}
plot(density(perfumes$Vanilla), col="blue", main="Vanilla and Floral density")
lines(density(perfumes$Floral), col="red")

legend(10, 0.2, legend=c("Vanilla", "Floral"),
       col=c("blue", "red"), lty=1:2, cex=0.8)

```

## PART B'

Now that we have an idea of the variable density, we can introduce the notion of position indicator. Let's take the variable `Vanilla` again. We now want to know which notes "cut" the population into two groups of equal size. The median is the value separating the higher half from the lower half of a data sample, a population, or a probability distribution. For a data set, it may be thought of as "the middle" value. To generalize, we have the notion of quantile, which are cut points dividing the observations in a sample in the same way. For example, the median is a quantile associate to the order 0.5. 

Lets illustrate : 
```{r}
d<-density(perfumes$Vanilla)
# Plot the line
plot(d, main="Vanilla Distribution and median")
p50=which.max(cumsum(d$y/sum(d$y)) >= 0.50)

# Plot the shading
polygon(c(-5, d$x[1:p50], d$x[p50]), c(0, d$y[1:p50], 0), col = 'lightblue')
```

We can do the same with the first and third quartile:
```{r}
d<-density(perfumes$Vanilla)

# Plot the line
plot(d, main="Vanilla Distribution and quartile")
p25=which.max(cumsum(d$y/sum(d$y)) >= 0.25)
p75=which.max(cumsum(d$y/sum(d$y)) >= 0.75)

# Plot the shading
polygon(c(-5, d$x[1:p25], d$x[p25]), c(0, d$y[1:p25], 0), col = 'lightblue')
poly_range <- d$x > d$x[p75] 
polygon(c(d$x[p75], d$x[poly_range], 15),c(0, d$y[poly_range], 0),col = "lightblue") 
```
Finaly, we plot the density with the mean and 2.5 and 97.5 quantile.
```{r}
d<-density(perfumes$Vanilla)

# Plot the line
plot(d, main="Vanilla Distribution and quantiles")
p025=which.max(cumsum(d$y/sum(d$y)) >= 0.025)
p975=which.max(cumsum(d$y/sum(d$y)) >= 0.975)

# Plot the shading
polygon(c(-5, d$x[1:p025], d$x[p025]), c(0, d$y[1:p025], 0), col = 'lightblue')
poly_range <- d$x > d$x[p975] 
polygon(c(d$x[p975], d$x[poly_range], 15),c(0, d$y[poly_range], 0),col = "lightblue")

# Plot the vline
abline(v=mean(perfumes$Vanilla))
text(mean(perfumes$Vanilla),0.2, "mean", pos=2)

abline(v=d$x[p975])
text(d$x[p975],0.2, "Q97.5", pos=2)

abline(v=d$x[p025])
text(d$x[p025],0.2, "Q0.25", pos=2)
```

## 1C

Dispersion indicator are useful to see how observations are distribute along the mean. 
Standard deviation is an indicator independant of units. It gives you an idea of how far you are to the mean. 

for 3 variables : `Vanilla`, `Floral`, `Citrus` we represent a tabular with their means, variance, median

|      | mean | variance | median | 
|      |-----------|-----------|-----------|
|Citrus| | | | | 
|Vanilla| | | | | 
|Floral| | | | | 

These position indicator are helpful to build another plot to represent the variable : the Boxplot.

```{r}
par(mfrow=c(1,3))
boxplot(perfumes$Floral, main="Floral")
text(0.75,mean(perfumes$Floral)+0.25, "mean", pos=2)
abline(h=mean(perfumes$Floral))
text(0.75,min(perfumes$Vanilla)+0.25, "min", pos=2)
abline(h=min(perfumes$Vanilla))
text(0.75,quantile(perfumes$Floral, probs = 0.25)+0.25, "q1", pos=2)
abline(h=quantile(perfumes$Floral, probs = 0.25))
text(0.75,quantile(perfumes$Floral, probs = 0.75)+0.25, "q3", pos=2)
abline(h=quantile(perfumes$Floral, probs = 0.75))
text(0.75,quantile(perfumes$Floral, probs = 0.5)+0.25, "Median", pos=2)
abline(h=quantile(perfumes$Floral, probs = 0.5))
text(0.75,quantile(perfumes$Floral, probs = 1)+0.25, "max", pos=2)
abline(h=quantile(perfumes$Floral, probs = 1))

boxplot(perfumes$Vanilla, main="Vanilla")
text(0.75,mean(perfumes$Vanilla)+0.25, "mean", pos=2)
abline(h=mean(perfumes$Vanilla))

boxplot(perfumes$Citrus, main="Citrus")
abline(h=mean(perfumes$Citrus))
text(0.75,mean(perfumes$Citrus)+0.25, "mean", pos=2)
```

# PARTIE 2

Here we introduce 2 packages : dyplr and ggplot to upgrate our graphics. We also considerate a product effect, as product influence the variance of the notes.

## 2A

```{r}
library(tidyverse)
```
As they are many products, we will focus on only three : 

```{r}
perfumes <- perfumes %>% filter(Product=="Angel"|Product=="Chanel N5"|Product=="J'adore ET")
```

```{r}
ggplot(perfumes)+aes(x=Floral, color=Product)+geom_density()
```
```{r}
ggplot(perfumes)+aes(y=Floral, x= Product, color=Product)+geom_boxplot()
```
# PART 3

To have a better understanding of this differences between products, we use the precedent boxplots, where we add a segment between each means. 
```{r}
ggplot(perfumes)+aes(y=Floral, x= Product, color=Product)+geom_boxplot() + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")
```
These segments are equivalent to a `distance` between all the product, in the `Floral` Point of view. If we want to have these visualization for all variables : 

```{r}
library("gridExtra")
A1 <- ggplot(perfumes)+aes(y=Floral, x= Product, color=Product)+geom_boxplot() + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
A2 <- ggplot(perfumes)+aes(y=Vanilla, x= Product, color=Product)+geom_boxplot() + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
A3 <- ggplot(perfumes)+aes(y=Citrus, x= Product, color=Product)+geom_boxplot() + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
grid.arrange(A1, A2, A3, ncol=2, nrow = 2)
```

without boxplots :
```{r}
library("gridExtra")
A1 <- ggplot(perfumes)+aes(y=Floral, x= Product, color=Product)+ 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
A2 <- ggplot(perfumes)+aes(y=Vanilla, x= Product, color=Product) + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
A3 <- ggplot(perfumes)+aes(y=Citrus, x= Product, color=Product) + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
grid.arrange(A1, A2, A3, ncol=2, nrow = 2)
```
without variance:
```{r}
perfumes$Vanilla <- perfumes$Vanilla/var(perfumes$Vanilla)
perfumes$Floral <- perfumes$Floral/var(perfumes$Floral)
perfumes$Citrus <- perfumes$Citrus/var(perfumes$Citrus)

library("gridExtra")

A1 <- ggplot(perfumes)+aes(y=Floral, x= Product, color=Product)+ 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
A2 <- ggplot(perfumes)+aes(y=Vanilla, x= Product, color=Product) + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
A3 <- ggplot(perfumes)+aes(y=Citrus, x= Product, color=Product) + 
stat_summary(fun=mean, geom="line", aes(group=1), color="black") + 
stat_summary(fun=mean, geom="point")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
grid.arrange(A1, A2, A3, ncol=2, nrow = 2)
```

We have finally 3 univariate analysis of the mean differences between all the product for each attributes. For example, Chanel N5 and J'adore ET are closer than Angel in Vanilla perception. They have a shorter "distance" between them. 

We can summary these graphics in 3 4x4 matrix with the distance mean-mean for each product : 
```{r}
Mean_Attributes <- perfumes %>% group_by(Product) %>% summarise(
          mean_Vanilla=mean(Vanilla),
          mean_Citrus=mean(Citrus),
          mean_Floral=mean(Floral)
)


matrix_Floral <- as.matrix(dist(Mean_Attributes$mean_Floral))
matrix_Citrus <- as.matrix(dist(Mean_Attributes$mean_Citrus))
matrix_Vanilla <- as.matrix(dist(Mean_Attributes$mean_Vanilla))

```

We can combine to visualize simultaneously 2 attributes:
```{r}
A1A2<-ggplot(Mean_Attributes)+aes(x = mean_Vanilla, y=mean_Floral, color=Product)+geom_path(aes(group=1),color="black")+geom_point()
A2A3<-ggplot(Mean_Attributes)+aes(x = mean_Floral, y=mean_Citrus, color=Product)+geom_path(aes(group=1),color="black")+geom_point()
A1A3<-ggplot(Mean_Attributes)+aes(x = mean_Vanilla, y=mean_Citrus, color=Product)+geom_path(aes(group=1), color="black")+geom_point()
grid.arrange(A1A2, A2A3, A1A3, ncol=2, nrow = 2)
```

Covariance and correlation notion : 
(Faire a la main?)
```{r}
cov(Mean_Attributes[,-1])
cov(perfumes[,c("Vanilla","Citrus", "Floral")])
```
```{r}
Mean_Attributes<- data.frame(Mean_Attributes, row.names=1)
```

with `Mean_Attributes`, we can construct two matrices : 
```{r}

Attributes_Mat <- cov(perfumes[,c("Vanilla","Citrus", "Floral")])
Products_Mat <- as.matrix(dist(Mean_Attributes))
t(Mean_Attributes)
```

```{r}
heatmap(Attributes_Mat)
heatmap(Products_Mat)
```


